---
title: Measure repeatability of sound segment classification
subtitle: Understanding cultural evolution in hummingbird leks through the fossilized birth-death process
author: <a href="https://marce10.github.io/">Marcelo Araya-Salas</a>
date: "`r Sys.Date()`"
toc: true
toc-depth: 3
toc-location: left
number-sections: true
highlight-style: pygments
format:
  html:
    df-print: kable
    code-fold: show
    code-tools: true
    css: qmd.css
editor_options: 
  chunk_output_type: console
---


```{r set root directory}
#| eval: true
#| echo: false

# install knitr package if not installed
if (!requireNamespace("knitr", quietly = TRUE)) {
  install.packages("knitr")
}

# set working directory 
knitr::opts_knit$set(root.dir =  "..")

```

```{r add link to github repo}
#| eval: true
#| output: asis
#| echo: false

# print link to github repo if any
if (file.exists("./.git/config")){
  config <- readLines("./.git/config")
  url <- grep("url",  config, value = TRUE)
  url <- gsub("\\turl = |.git$", "", url)
  if (nchar(url) > 1)
  cat("\nSource code and data found at [", url, "](", url, ")", sep = "")
  }

```

```{r setup style}
#| message: false
#| warning: false

# options to customize chunk outputs
knitr::opts_chunk$set(
  tidy.opts = list(width.cutoff = 65), 
  tidy = TRUE,
  message = FALSE
 )

```

<!-- skyblue box -->

::: {.alert .alert-info}
# Purpose {.unnumbered .unlisted}

- Measure repeatability of segment classification between two observers using Cohen's kappa statistic

:::


# Load packages {.unnumbered .unlisted}

```{r load packages}

library("irr")

```

Measure repeatability of segment labels

Inter-observer reliability:
```{r}
segments_2_observers <- read.csv(
  "./data/processed/segment_classification_2_observers.csv"
)
```

` length(unique(segments_2_observers$lek.song.type))` segments classified by two observers

```{r}
kp <- irr::kappa2(segments_2_observers[,
  c("obs1", "obs2")
])

kp

```

::: {.alert .alert-success}
# Takeaways {#takeaways .unnumbered .unlisted}

High repeatability of segment classification between two observers, with a Cohen's kappa of `r round(kp$value, 3)`, indicating substantial agreement.

:::

------------------------------------------------------------------------


<!-- add packages used, system details and versions  -->

::: {.callout-note collapse="true"}
## Session information {.unnumbered .unlisted}

```{r session info}
#| echo: false

# if devtools is installed use devtools::session_info()
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::session_info()
} else {
  sessionInfo()
}

```
:::
