---
title: Measure song similarity 
subtitle: Understanding cultural evolution of songs in hummingbird leks through the fossilized birth-death process
author: <a href="https://marce10.github.io/">Marcelo Araya-Salas</a>
date: "`r Sys.Date()`"
toc: true
toc-depth: 3
toc-location: left
number-sections: true
highlight-style: pygments
format:
  html:
    df-print: kable
    code-fold: show
    code-tools: true
    css: qmd.css
editor_options: 
  chunk_output_type: console
---


```{r set root directory}
#| eval: true
#| echo: false

# install knitr package if not installed
if (!requireNamespace("knitr", quietly = TRUE)) {
  install.packages("knitr")
}

# set working directory 
knitr::opts_knit$set(root.dir =  "..")

```

```{r add link to github repo}
#| eval: true
#| output: asis
#| echo: false

# print link to github repo if any
if (file.exists("./.git/config")){
  config <- readLines("./.git/config")
  url <- grep("url",  config, value = TRUE)
  url <- gsub("\\turl = |.git$", "", url)
  if (nchar(url) > 1)
  cat("\nSource code and data found at [", url, "](", url, ")", sep = "")
  }

```

```{r setup style}
#| message: false
#| warning: false

# options to customize chunk outputs
knitr::opts_chunk$set(
  tidy.opts = list(width.cutoff = 65), 
  tidy = TRUE,
  message = FALSE
 )

```

<!-- skyblue box -->

::: {.alert .alert-info}
# Purpose {.unnumbered .unlisted}

- Evaluate the effect of lek membership in song similarity 

:::


# Load packages {.unnumbered .unlisted}

```{r load packages}

library("TraMineR")
library("brms")
library("brmsish")

```

## Statistical modeling

- Pairwise song similarity was analyzed using a **Bayesian Gaussian linear mixed model**.

- The response variable was the **pairwise Optimal Matching (OM) distance** between song sequences.

- Pairwise distances were calculated using **Optimal Matching** with biologically informed substitution costs that assigned lower penalties to substitutions within broad acoustic categories (pure tones or trills) than between categories.

- Analyses were restricted to **pairwise comparisons between songs recorded within the same study site**, allowing us to test whether songs were more similar within leks than between neighboring leks while minimizing large-scale geographic effects.

- Candidate fixed effects included:
  - `same_lek`, indicating whether two songs originated from the same lek,
  - `year_diff`, representing the minimum temporal separation (years) between recorded occurrences of the two song types,
  - and the interaction `same_lek × year_diff`.

- Because each song type appeared in multiple pairwise comparisons, song identity was included as a **multi-membership random intercept** (`mm(song1, song2)`) to account for the non-independence of pairwise observations.

- The fitted model was:

```r
distance ~ same_lek * year_diff +
  (1 | mm(song1, song2))
```

- Models were fitted in **Stan** through the R package **brms** using a Gaussian likelihood and weakly informative default priors.

- Model adequacy was assessed using posterior predictive checks.

```{r}
# Read data

dat <- read.csv("./data/raw/segments_by_song_type.csv")

# Create sequence object

# Columns containing the 20 song segments
seqs <- dat[, grep("^X", names(dat))]

# Create sequence object
seq.obj <- seqdef(seqs)

# Compute Optimal Matching distances

states <- c("d","p","u","s","m","f")

sm <- matrix(2,
             nrow = 6,
             ncol = 6,
             dimnames = list(states, states))

diag(sm) <- 0

# within pure tones
sm["d","p"] <- sm["p","d"] <- 1
sm["p","u"] <- sm["u","p"] <- 1
sm["d","u"] <- sm["u","d"] <- 1

# within trills
sm["s","m"] <- sm["m","s"] <- 1
sm["m","f"] <- sm["f","m"] <- 1
sm["s","f"] <- sm["f","s"] <- 1

dist.mat <- seqdist(
  seq.obj,
  method = "OM",
  sm = sm,
  indel = 3,
  norm = "maxlength"
)
# Convert distance matrix to pairwise data frame
pairs <- which(upper.tri(dist.mat), arr.ind = TRUE)

pair.dat <- data.frame(
  song1    = dat$song.type[pairs[,1]],
  song2    = dat$song.type[pairs[,2]],
  lek1     = dat$lek[pairs[,1]],
  lek2     = dat$lek[pairs[,2]],
  distance = dist.mat[pairs]
)

# Predictor variables
years <-
  read.csv("./data/raw/year_range_by_song_type.csv", stringsAsFactors = FALSE)


pair.dat$min.year.1 <- sapply(pair.dat$song1, function(x) {
  min(years$year[years$song.type == x])
})

pair.dat$max.year.1 <- sapply(pair.dat$song1, function(x) {
  max(years$year[years$song.type == x])
})

pair.dat$min.year.2 <- sapply(pair.dat$song2, function(x) {
  min(years$year[years$song.type == x])
})

pair.dat$max.year.2 <- sapply(pair.dat$song2, function(x) {
  max(years$year[years$song.type == x])
})


# Merge with pairwise data
pair.dat$year.diff <- sapply(1:nrow(pair.dat), function(i) {
  song1 <- pair.dat$song1[i]
  song2 <- pair.dat$song2[i]
  year1 <- years$year[years$song.type == song1]
  year2 <- years$year[years$song.type == song2]
 
  # min of all possible differences
  min(outer(year1, c(year2, 2222), FUN = function(x, y) abs(x - y)))
  
})


lek.site <- data.frame(
  lek = c(
    "BR1", "BR2",
    "CCE", "CCL", "LOC", "SAT", "SJA", "STR", "SUR",
    "TR1", "TR2",
    "HC1", "HC2"
  ),
  site = c(
    rep("BR", 2),
    rep("LS", 7),
    rep("TR", 2),
    rep("HC", 2)
  ),
  stringsAsFactors = FALSE
)

pair.dat$same_site <- sapply(1:nrow(pair.dat), function(i) {
  lek1 <- pair.dat$lek1[i]
  lek2 <- pair.dat$lek2[i]
  site1 <- lek.site$site[lek.site$lek == lek1]
  site2 <- lek.site$site[lek.site$lek == lek2]
  site1 == site2
})

# keep only from same site
pair.dat <- pair.dat[pair.dat$same_site, ]


# head(pair.dat)

# Songs from the same lek?
pair.dat$same_lek <- pair.dat$lek1 == pair.dat$lek2

# Inspect the distribution of distances

hist(pair.dat$distance, breaks = 30)

# Bayesian regression

fit <- brm(
  distance ~ same_lek * year.diff +               
    (1 | mm(song1, song2)),
  data = pair.dat,
  family = gaussian(),
  chains = 4,
  cores = 4,
  iter = 5000,
  control = list(
    adapt_delta = 0.99,
    max_treedepth = 15
  ),
  file = "./data/processed/song_similarity_model.rds", 
  file_refit = "on_change"
)

# Summarize results

summary(fit)

```


```{r}
###############################################################
# First year of each song type
###############################################################

song.info <- unique(rbind(
  data.frame(song = pair.dat$song1,
             lek = pair.dat$lek1,
             min.year = pair.dat$min.year.1),
  data.frame(song = pair.dat$song2,
             lek = pair.dat$lek2,
             min.year = pair.dat$min.year.2)
))

###############################################################
# Add same_lek if needed
###############################################################

pair.dat$same_lek <- pair.dat$lek1 == pair.dat$lek2

###############################################################
# Nearest neighbour within one year before first occurrence
###############################################################

songs <- song.info$song

nearest.same.lek <- rep(NA, length(songs))

for(i in seq_along(songs)){

  song <- songs[i]

  t0 <- song.info$min.year[song.info$song == song]

  # all pairwise comparisons involving this song
  x <- pair.dat[pair.dat$song1 == song | pair.dat$song2 == song, ]

  if(nrow(x) == 0)
    next

# First year of the other song
other.year <- ifelse(
  x$song1 == song,
  x$min.year.2,
  x$min.year.1
)

# Candidate songs:
# Candidate songs:
# first appeared in the same year or exactly one year earlier
keep <- other.year %in% c(t0 - 1, t0)

x <- x[keep, ]

if(nrow(x) == 0)
  next

# Was there at least one candidate from the same lek?
if(!any(x$same_lek)){
  nearest.same.lek[i] <- NA
  next
}

  # nearest neighbour
  mindist <- min(x$distance)
  closest <- x[x$distance == mindist, ]

  nearest.same.lek[i] <- any(closest$same_lek)
}

table(nearest.same.lek, useNA = "ifany")

mean(nearest.same.lek, na.rm = TRUE)


sum(nearest.same.lek, na.rm = TRUE) / sum(!is.na(nearest.same.lek))


```

::: {.alert .alert-info}

## Results

- Posterior predictive checks indicated that the Gaussian model adequately described the observed distribution of pairwise song distances.

```{r}
pp_check(fit)
```


- Songs recorded within the **same lek** were consistently more similar than songs recorded in **different leks within the same study site** (posterior estimate = -0.031).

```{r}
fixef(fit)
```

- Pairwise song distance increased with the temporal separation between recordings for **between-lek comparisons**, indicating that songs became progressively more dissimilar through time even among neighboring leks.

```{r}

conditional_effects(fit, effects = "year.diff:same_lek")

hypothesis(
  fit,
  "year.diff > 0"
)
```

- The temporal increase in song dissimilarity was **significantly stronger within leks** than between leks, as indicated by a positive interaction between `same_lek` and `year_diff`.

```{r}
hypothesis(
  fit,
  "same_lekTRUE:year.diff > 0"
)
```

- Consequently, pairwise song dissimilarity increased through time **within leks**, with overwhelming posterior support for a positive temporal trend.

```{r}
hypothesis(
  fit,
  "year.diff + same_lekTRUE:year.diff > 0"
)
```

::: 

::: {.alert .alert-success}
# Takeaways {#takeaways .unnumbered .unlisted}

- Together, these results indicate that songs exhibit strong local cultural structure, with song types being more similar within leks than between neighboring leks, while simultaneously accumulating cultural differences through time. Moreover, the rate at which song dissimilarity accumulated was greater within leks than between leks.
:::

------------------------------------------------------------------------


<!-- add packages used, system details and versions  -->

::: {.callout-note collapse="true"}
## Session information {.unnumbered .unlisted}

```{r session info}
#| echo: false

# if devtools is installed use devtools::session_info()
if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::session_info()
} else {
  sessionInfo()
}

```
:::
