---
title: <font size="6"><b>Tree topology diagnostics</b></font>
subtitle: Long-billed hermit song cultural evolution
author: <font size="4"><a href="https://marce10.github.io/">Marcelo Araya-Salas PhD</a></font>
date: "`r Sys.Date()`"
toc: true
toc-depth: 5
toc-location: left
number-sections: true
highlight-style: pygments
format:
  html:
    df-print: kable
    code-fold: show
    code-tools: true
    css: qmd.css
editor_options: 
  chunk_output_type: console
---

```{r set root directory}
#| eval: true
#| echo: false

# set working directory 
knitr::opts_knit$set(root.dir =  "..")

```

```{r add link to github repo}
#| eval: true
#| output: asis
#| echo: false

# print link to github repo if any
if (file.exists("./.git/config")){
  config <- readLines("./.git/config")
  url <- grep("url",  config, value = TRUE)
  url <- gsub("\\turl = |.git$", "", url)
  cat("\nSource code and data found at [", url, "](", url, ")", sep = "")
  }

```

## Load packages {.unnumbered .unlisted}
```{r clean session}
#| eval: true
#| echo: false
#| warning: false
#| message: false


## vector with package names
x <- c("pbapply", "viridis", "knitr", "kableExtra", "ggplot2", "ape", "PhenotypeSpace", "brms", "brmsish", "cowplot", "ggridges")

aa <- lapply(x, function(y) {
  
  # check if installed, if not then install 
  if (!y %in% installed.packages()[,"Package"]) 
    install.packages(y) 

  # load package
  try(require(y, character.only = T), silent = T)
})

```

## Custom functions and global options {.unnumbered .unlisted}
```{r functions and parameters}
#| eval: true
#| 
knitr::opts_chunk$set(out.width = "100%", out.height = "100%", echo = TRUE) 

source('~/Dropbox/Projects/lbh_cultural_evolution_revBayes/source/custom_treespace.R')

theme_set(theme_classic(base_size = 10))

fill_color <- viridis::mako(10, alpha = 0.5)[7]

iterations <- 5000
chains <- 4
priors <- c(set_prior("normal(0, 1.5)", class = "Intercept"), set_prior("normal(0, 1.5)", class = "b"))

# to create several posterior predictive check plots out of a brms fit
custom_ppc1 <- function(fit, group = NULL, ndraws = 30) {
  
    ppc_dens <-
        pp_check(fit,
                 ndraws = ndraws,
                 type = 'dens_overlay_grouped',
                 group = group)  # shows dens_overlay plot by default
    pp_mean <-
        pp_check(
            fit,
            type = "stat_grouped",
            stat = "mean",
            group = group,
            ndraws = ndraws
        )
    pp_scat <-
        pp_check(fit,
                 type = "scatter_avg_grouped",
                 group = group,
                 ndraws = ndraws)
    pp_stat2 <- pp_check(fit, type = "stat_2d", ndraws = ndraws)
    pp_loo <- pp_check(fit, type = "loo_pit_qq", ndraws = ndraws)
    pp_error <-
        pp_check(fit,
                 type = "error_scatter_avg_grouped",
                 ndraws = ndraws,
                 group = group)
    plot_l <-
        list(ppc_dens, pp_mean, pp_scat, pp_error,  pp_stat2, pp_loo)
    
    plot_l[c(1, 3:length(plot_l))] <-
        lapply(plot_l[c(1, 3:length(plot_l))], function(x)
            x  + scale_color_viridis_d(
                begin = 0.1,
                end = 0.8,
                alpha = 0.5
            ) + scale_fill_viridis_d(
                begin = 0.1,
                end = 0.8,
                alpha = 0.5
            ) + theme_classic())     

    print(plot_grid(plotlist = plot_l, ncol = 2))
}

custom_ppc <- function(...) suppressMessages(custom_ppc1(...))

```

## Read data {.unnumbered .unlisted}
```{r, eval = FALSE, echo = FALSE}

# read revbayes output 
rb.trees <- load.multi("~/Dropbox/Projects/lbh_cultural_evolution/output/most_recent_revbayes_models/", format = "revbayes")

saveRDS(rb.trees, "./output/revbayes_output_in_single_R_object.RDS")

```


```{r}
#| eval: false

rb.trees <- readRDS("./output/revbayes_output_in_single_R_object.RDS")

rb.trees <- rb.trees[grep("run_", names(rb.trees), invert = TRUE, value = TRUE)]

trees_diags <- data.frame(model = names(rb.trees))

# get lek name
trees_diags$lek <- sapply(strsplit(trees_diags$model, "_", fixed = TRUE), "[", 1)

# substitution model
trees_diags$subs <- sapply(strsplit(trees_diags$model, "_", fixed = TRUE), "[", 2)

# period
trees_diags$period <- sapply(strsplit(trees_diags$model, "_", fixed = TRUE), "[", 3)

# and which fossils were used
trees_diags$fossils <- sapply(strsplit(trees_diags$model, "_", fixed = TRUE), "[", 4)

trees_diags$align <- sapply(strsplit(trees_diags$model, "_", fixed = TRUE), "[", 5)

trees_diags$n_trees <- sapply(rb.trees, function(x) length(x$trees))

# Number of models by iterations and lek
kbl <-
    knitr::kable(as.matrix(table(trees_diags$lek, trees_diags$n_trees)), caption = "Number of models with a specific number of trees by lek")

kableExtra::kable_styling(kbl)

# Number of models by iterations and lek
kbl <-
    knitr::kable(as.matrix(table(trees_diags$lek, trees_diags$subs)), caption = "Number of models with a specific number of trees by lek")

kableExtra::kable_styling(kbl)


```

# Topological space estimation

* Pairwise topological distance between all trees for each lek
* 100 trees for each model evenly spaced along the chain
* Comparing based on shared song types (pairwise shared tips)
* Some trees were removed if NAs were produced when comparing topologies (i.e. non-comparable topologies)

## All fossils
```{r make tree distances all fossils, eval = FALSE}

#selected leks
sel_leks <- c("SUR", "CCE", "HC1", "BR1", "TR1")

tree_names <- grep("early", names(rb.trees), invert = TRUE, value = TRUE)

pnts_lks <- lapply(sel_leks, function(i) {
  
  print(i)
  lek_trees <- grep(i, tree_names, value = TRUE)
  
  multi_list <- lapply(lek_trees, function(x) read.tree(file.path("./output/most_recent_revbayes_models/", x), keep.multi = TRUE))
  
  names(multi_list) <- sapply(basename(lek_trees), function(x) paste(strsplit(x, "_", fixed = TRUE)[[1]][1:5], collapse = "_"))
  
  pnts <- try(custom_treespace(multi_list, n.points = 100, method = "RF", cl = 10), silent = TRUE)

  if (!is(pnts, "try-error")) 
    return(pnts) else
      return(NULL)
})

names(pnts_lks) <- sel_leks

sapply(pnts_lks, is.null)

saveRDS(pnts_lks, file = "./output/topology/tree_distance_all_fossils_pairwise_shared.RDS")


```

## Early fossils
```{r make tree distances early fossils, eval = FALSE}

#selected leks
sel_leks <- c("SUR", "CCE", "HC1", "BR1", "TR1")
# sel_leks <- c("SUR", "TR1")

tree_names <- grep("early", names(rb.trees), value = TRUE)

pnts_lks <- lapply(sel_leks, function(i) {
  
  print(i)
  lek_trees <- grep(i, tree_names, value = TRUE)
  
  multi_list <- lapply(lek_trees, function(x) read.tree(file.path("./output/most_recent_revbayes_models/", x), keep.multi = TRUE))
  
  names(multi_list) <- sapply(basename(lek_trees), function(x) paste(strsplit(x, "_", fixed = TRUE)[[1]][1:5], collapse = "_"))
  
  pnts <- try(custom_treespace(multi_list, n.points = 100, method = "RF", cl = 10), silent = TRUE)

  if (!is(pnts, "try-error")) 
    return(pnts) else
      return(NULL)
})

names(pnts_lks) <- sel_leks

sapply(pnts_lks, is.null)

saveRDS(pnts_lks, file = "./output/topology/tree_distance_early_fossils_pairwise_shared.RDS")

```

# Plot topological spaces

## All fossils

### New and old data sets
```{r plot tree distances ggplots all fossils, eval = TRUE, warning=FALSE}

pnts_lks <- readRDS("./output/topology/tree_distance_all_fossils_pairwise_shared.RDS")

ggs <- lapply(pnts_lks, function(X) {

    ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, ncol = 2) +
      scale_fill_gradientn(colours = mako(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") + 
      ggtitle(label = sprintf("Tree space for %d trees", nrow(X) / length(unique(X$chain)))) 
    })

ggs
```

### Old data set
```{r plot tree distances ggplots old all fossils, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("old", X$chain), ]   
  
  if (nrow(X) > 0)
  ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = mako(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain))))
    })

ggs
```

### New data set
```{r plot tree distances ggplots new all fossils, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("new", X$chain), ]   
  
  ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = mako(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %d trees", nrow(X) / length(unique(X$chain))))

    })

ggs
```

#### Clumping molecular clocks

##### Old data set
```{r plot tree distances ggplots old all fossils clumped clocks, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("old", X$chain), ]   
  X$clock <- ifelse(grepl("global", X$chain), "Global", "Uexp")
  X$chain <- gsub("_global|_Uexp", "", X$chain)
  
    if (nrow(X) > 0)
    ggplot(data = X, aes(x = x, y = y)) +
    geom_path(alpha = 0.1, size = 0.75, show.legend = FALSE) +
    geom_point(aes(shape = clock,  col = clock), size = 4) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_color_viridis_d(option = "G", begin = 0.2, end = 0.8, alpha = 0.75) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %d trees", nrow(X) / length(unique(X$chain))))

    })

ggs
```

##### New data set
```{r plot tree distances ggplots new all fossils clumped clocks, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("new", X$chain), ]   
  X$clock <- ifelse(grepl("global", X$chain), "Global", "Uexp")
  X$chain <- gsub("_global|_Uexp", "", X$chain)
  
  ggplot(data = X, aes(x = x, y = y)) +
    geom_path(alpha = 0.1, size = 0.75, show.legend = FALSE) +
    geom_point(aes(shape = clock,  col = clock), size = 4) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_color_viridis_d(option = "G", begin = 0.2, end = 0.8, alpha = 0.75) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %d trees", nrow(X) / length(unique(X$chain))))

    })

ggs
```


## Early fossils

### New and old data sets
```{r plot tree distances ggplots early fossils, eval = TRUE, warning=FALSE}

pnts_lks <- readRDS("./output/topology/tree_distance_early_fossils_pairwise_shared.RDS")

ggs <- lapply(pnts_lks, function(X) {

    ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = mako(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") + 
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain)))) 

    })

ggs

```

### New data set
```{r plot tree distances ggplots new early fossils, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("new", X$chain), ]   
  
  ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = mako(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain))))
  })

ggs


```

### Old data set
```{r plot tree distances ggplots old early fossils, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("old", X$chain), ]   
  
  if (nrow(X) > 0)
  ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = mako(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain))))
  })

ggs


```

### Clumping molecular clocks

#### Old data set
```{r plot tree distances ggplots old early fossils clumped clocks, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("old", X$chain), ]   
  X$clock <- ifelse(grepl("global", X$chain), "Global", "Uexp")
  X$chain <- gsub("_global|_Uexp", "", X$chain)
  
    if (nrow(X) > 0)
    ggplot(data = X, aes(x = x, y = y)) +
    geom_path(alpha = 0.1, size = 0.75, show.legend = FALSE) +
    geom_point(aes(shape = clock,  col = clock), size = 4) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_color_viridis_d(option = "G", begin = 0.2, end = 0.8, alpha = 0.75) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain))))

    })

ggs
```

#### New data set
```{r plot tree distances ggplots new early fossils clumped clocks, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("new", X$chain), ]   
  X$clock <- ifelse(grepl("global", X$chain), "Global", "Uexp")
  X$chain <- gsub("_global|_Uexp", "", X$chain)
  
  ggplot(data = X, aes(x = x, y = y)) +
    geom_path(alpha = 0.1, size = 0.75, show.legend = FALSE) +
    geom_point(aes(shape = clock,  col = clock), size = 4) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_color_viridis_d(option = "G", begin = 0.2, end = 0.8, alpha = 0.75) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain))))

    })

ggs
```

# Topological space dissimilarity

## Observations per lek and alignment

All fossils
```{r, eval = TRUE, message=FALSE, warning=FALSE}
# similarity
out <- lapply(pnts_lks, function(x){
  
  Y <- space_similarity(formula = chain ~ x + y, data = x, pb = FALSE, type = "density.overlap")
  
  Y$mean.overlap <- Y$mean.overlap #-  mean(Y$mean.overlap)
  # Y$overlap <- Y$overlap / max(Y$overlap)
  
  return(Y)  
})

topo_similarity_all <- do.call(rbind, out)

topo_similarity_all2 <- topo_similarity_all[, c("group.1",  "group.2", "overlap.1in2")]
names(topo_similarity_all2) <- c("group.1", "group.2", "mean.overlap")

topo_similarity_all3 <- topo_similarity_all[, c("group.2",  "group.1", "overlap.2in1")]
names(topo_similarity_all3) <- c("group.2", "group.1", "mean.overlap")

topo_similarity_all <- rbind(topo_similarity_all2, topo_similarity_all3)

topo_similarity_all$alignment <- sapply(topo_similarity_all$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][1])

topo_similarity_all$data.set <- sapply(topo_similarity_all$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][2])

topo_similarity_all$fossils <- sapply(topo_similarity_all$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][3])

topo_similarity_all$models <- sapply(topo_similarity_all$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][4])


topo_similarity_all$lek <- substr(rownames(topo_similarity_all), 0 , 3)

tab <- as.data.frame(table(topo_similarity_all$lek, topo_similarity_all$alignment))
names(tab) <- c("lek", "model", "count")

tab
```

Early fossils
```{r}
out <- lapply(pnts_lks, function(x){
  
  Y <- space_similarity(formula = chain ~ x + y, data = x, pb = FALSE, type = "mean.density.overlap")
  
  Y$mean.overlap <- Y$mean.overlap -  mean(Y$mean.overlap)
  # Y$overlap <- Y$overlap / max(Y$overlap)
  
  return(Y)  
})

topo_similarity_early <- do.call(rbind, out)

topo_similarity_early2 <- topo_similarity_early[, c("group.1",  "group.2", "overlap.2in1")]
names(topo_similarity_early2) <- c("group.2", "group.1", "mean.overlap")

topo_similarity_early3 <- topo_similarity_early[, c("group.1",  "group.2", "overlap.1in2")]
names(topo_similarity_early3) <- c("group.2", "group.1", "mean.overlap")

topo_similarity_early <- rbind(topo_similarity_early2, topo_similarity_early3)

topo_similarity_early$alignment <- sapply(topo_similarity_early$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][1])

topo_similarity_early$data.set <- sapply(topo_similarity_early$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][2])

topo_similarity_early$fossils <- sapply(topo_similarity_early$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][3])

topo_similarity_early$models <- sapply(topo_similarity_early$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][4])

topo_similarity_early$lek <- substr(rownames(topo_similarity_early), 0 , 3)

tab <- as.data.frame(table(topo_similarity_early$lek, topo_similarity_early$alignment))
names(tab) <- c("lek", "model", "count")

tab
```


## Plot dissimilarity by treatment
```{r}
#| eval: true
#| warning: false
#| out.height: '170%'

 

topo_similarity_all$fossils <- "all"
topo_similarity_early$fossils <- "early"

shared_columns <- intersect(names(topo_similarity_all), names(topo_similarity_early))

topo_similarity <- rbind(topo_similarity_all[, shared_columns], topo_similarity_early[, shared_columns])

topo_similarity$dissimilarity <- 1 - topo_similarity$mean.overlap

# raincloud plot:
ggplot(topo_similarity,
       aes(x = alignment, y = dissimilarity, fill = alignment)) +
  # add half-violin from {ggdist} package
  ggdist::stat_halfeye(
    # fill = fill_color,
    alpha = 0.5,
    # custom bandwidth
    adjust = .5,
    # adjust height
    width = .6,
    .width = 0,
    # move geom to the cright
    justification = -.2,
    point_colour = NA
  ) +
  geom_boxplot(# fill = fill_color,
    width = .15,
    # remove outliers
    outlier.shape = NA) +
    # add justified jitter from the {gghalves} package
    gghalves::geom_half_point(
      # color = fill_color,
      # draw jitter on the left
      side = "l",
      # control range of jitter
      range_scale = .4,
      # add some transparency
      alpha = .5,
      transformation = ggplot2::position_jitter(height = 0)
    ) +
      scale_fill_viridis_d(option = "G",
                           end = 0.8,
                           alpha = 0.6) +
           scale_x_discrete(labels = c(
        "natural" = "Natural",
        "artificial" = "Artificial",
        "simulated" = "Simulated"
      )) +
      theme(legend.position = "none") +
      facet_wrap(~ data.set + models + fossils, ncol = 2) +
  labs(x = "Alignment", y = "Topologic space dissimilarity") 

```

## Stats

### Dissimilarity of the topological space
```{r, eval = FALSE}

# topo_similarity$dissimilarity[topo_similarity$dissimilarity <= 0] <- 0.001 

topo_similarity$alignment <- factor(topo_similarity$alignment, levels = c("all.equal", "optimal", "prank"))

topo_similarity$data.set <- factor(topo_similarity$data.set, levels = c("new", "old"))

topo_similarity$models <- factor(topo_similarity$models, levels = c("Uexp", "global"))

topo_similarity$fossils <- factor(topo_similarity$fossils, levels = c("early", "all"))

mod <- brm(
  formula = dissimilarity ~ alignment + data.set + models + fossils + (1 | lek), 
  data = topo_similarity,
  iter = iterations,
  family = gaussian(),
  chains = chains,
  cores = chains,
   control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "./data/processed/regression_results_topological_similiarity",
  file_refit = "always")

mod <- add_criterion(mod, criterion = c("loo"))

null_mod <- brm(
  formula = dissimilarity ~ 1, 
  data = topo_similarity,
  iter = iterations,
  family = gaussian(),
  chains = chains,
  cores = chains,
   control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "./data/processed/null_regression_results_topological_similiarity",
  file_refit = "always"
)

null_mod <- add_criterion(null_mod, criterion = c("loo"))

```

Model selection:
```{r}
#| results: asis

mod <- readRDS("./data/processed/regression_results_topological_similiarity.rds")

null_mod <- readRDS("./data/processed/null_regression_results_topological_similiarity.rds")

loo_diff <- loo::loo_compare(mod, null_mod)

as.data.frame(loo_diff[,1:2], row.names = c("model", "null model"))
```

```{r}
#| eval: true
#| results: asis

extended_summary(fit = mod,
    n.posterior = 1000,
    fill =  mako(10)[7],
    trace.palette = viridis,
    remove.intercepts = TRUE,
    highlight = TRUE,
    trace = TRUE,
    return = FALSE, 
    print.name = FALSE
)


```

Contrasts for alignment strategies:
```{r}
#| eval: true
#| results: asis

# contrasts
contrasts(
    fit = mod,
    sort.levels = c("prank", "optimal", "all.equal"),
    predictor = "alignment",
    n.posterior = 2000,
    level.sep = " VS ",
    html.table = TRUE,
    plot = TRUE,
    highlight = TRUE,
    fill = mako(10)[7]
)

```

```{r}
#| warning: false

custom_ppc(fit = mod, group = "alignment")

```

# Session information {.unnumbered .unlisted}

```{r session info, echo=F}

sessionInfo()

```
