---
title: <center><font size="6"><b>Tree topology diagnostics</b></font></center>
subtitle: <center><font size="4"><b><a href="https://github.com/maRce10/lbh_cultural_evolution">Long-billed hermit song cultural evolution<a></b></font></center>
author: <center><font size="4"><a href="https://marce10.github.io/">Marcelo Araya-Salas PhD</a> & <a href="https://scholar.google.com/citations?user=0a8k9T8AAAAJ&hl=es&oi=ao"> Beatriz Willink PhD</a></font></center>
date: <center>`r format(Sys.Date(), "%d-%m-%Y")`</center>
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scroll: yes
fontsize: 12pt 
editor_options: 
  chunk_output_type: console
---

```{r clean session, eval = TRUE, echo = TRUE, message = FALSE, warning = FALSE}

## vector with package names
x <- c("pbapply", "viridis", "knitr", "kableExtra", "ggplot2", "ape", "PhenotypeSpace", "lmerTest", "brms", "brmsish")

aa <- lapply(x, function(y) {
  
  # check if installed, if not then install 
  if (!y %in% installed.packages()[,"Package"]) 
    install.packages(y) 

  # load package
  try(require(y, character.only = T), silent = T)
})

```

```{r functions and parameters, eval = TRUE}

knitr::opts_knit$set(root.dir = normalizePath(".."))

knitr::opts_chunk$set(dpi = 38, fig.width = 18, fig.height = 10, echo = TRUE) 

options(knitr.kable.NA = '')

source('~/Dropbox/Projects/lbh_cultural_evolution_revBayes/source/custom_treespace.R')

theme_set(theme_classic(base_size = 22))

fill_color <- viridis::mako(10, alpha = 0.5)[7]

```

```{r, eval = FALSE, echo = FALSE}

# read revbayes output 
rb.trees <- load.multi("~/Dropbox/Projects/lbh_cultural_evolution/output/most_recent_revbayes_models/", format = "revbayes")

saveRDS(rb.trees, "./output/revbayes_output_in_single_R_object.RDS")

```

```{r, eval = FALSE, echo = TRUE}

rb.trees <- readRDS("./output/revbayes_output_in_single_R_object.RDS")

rb.trees <- rb.trees[grep("run_", names(rb.trees), invert = TRUE, value = TRUE)]

trees_diags <- data.frame(model = names(rb.trees))

# get lek name
trees_diags$lek <- sapply(strsplit(trees_diags$model, "_", fixed = TRUE), "[", 1)

# substitution model
trees_diags$subs <- sapply(strsplit(trees_diags$model, "_", fixed = TRUE), "[", 2)

# period
trees_diags$period <- sapply(strsplit(trees_diags$model, "_", fixed = TRUE), "[", 3)

# and which fossils were used
trees_diags$fossils <- sapply(strsplit(trees_diags$model, "_", fixed = TRUE), "[", 4)

trees_diags$align <- sapply(strsplit(trees_diags$model, "_", fixed = TRUE), "[", 5)

trees_diags$n_trees <- sapply(rb.trees, function(x) length(x$trees))

# Number of models by iterations and lek
kbl <- knitr::kable(as.matrix(table(trees_diags$n_trees, trees_diags$lek)), caption = "Number of models with a specific number of trees by lek")

kableExtra::kable_styling(kbl)

```

---

# Tree spaces by leks

## All fossils using pairwise shared tips

* 100 trees for each model evenly spaced along the chain
* Some trees were removed if NAs were produced when comparing topologies (i.e. non-comparable topologies)

```{r make tree distances all fossils, eval = FALSE}

#selected leks
sel_leks <- c("SUR", "CCE", "HC1", "BR1", "TR1")

tree_names <- grep("early", names(rb.trees), invert = TRUE, value = TRUE)

pnts_lks <- lapply(sel_leks, function(i) {
  
  print(i)
  lek_trees <- grep(i, tree_names, value = TRUE)
  
  multi_list <- lapply(lek_trees, function(x) read.tree(file.path("./output/most_recent_revbayes_models/", x), keep.multi = TRUE))
  
  names(multi_list) <- sapply(basename(lek_trees), function(x) paste(strsplit(x, "_", fixed = TRUE)[[1]][1:5], collapse = "_"))
  
  pnts <- try(custom_treespace(multi_list, n.points = 100, method = "RF", cl = 10), silent = TRUE)

  if (!is(pnts, "try-error")) 
    return(pnts) else
      return(NULL)
})

names(pnts_lks) <- sel_leks

sapply(pnts_lks, is.null)

saveRDS(pnts_lks, file = "./output/topology/tree_distance_all_fossils_pairwise_shared.RDS")

```

#### New and old data sets
```{r plot tree distances ggplots all fossils, eval = TRUE, warning=FALSE}

pnts_lks <- readRDS("./output/topology/tree_distance_all_fossils_pairwise_shared.RDS")

ggs <- lapply(pnts_lks, function(X) {

    ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = viridis(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") + 
      ggtitle(label = sprintf("Tree space for %d trees", nrow(X) / length(unique(X$chain)))) 
    })

ggs
```

#### Old data set
```{r plot tree distances ggplots old all fossils, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("old", X$chain), ]   
  
  if (nrow(X) > 0)
  ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = viridis(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain))))
    })

ggs
```

#### New data set
```{r plot tree distances ggplots new all fossils, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("new", X$chain), ]   
  
  ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = viridis(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %d trees", nrow(X) / length(unique(X$chain))))

    })

ggs
```

#### Summary boxplots

Values were mean-centered within lek
```{r, eval = TRUE, message=FALSE, warning=FALSE}

out <- lapply(pnts_lks, function(x){
  
  Y <- space_size(formula = chain ~ x + y, data = x, pb = FALSE)
  Y$mean.c.size <- Y$size - mean(Y$size)
  # Y$size <- Y$size / max(Y$size)
 
  return(Y)  
})

topo_size_all <- do.call(rbind, out)

topo_size_all$alignment <- sapply(topo_size_all$group, function(x) strsplit(x, "_", fixed = TRUE)[[1]][1])

topo_size_all$data.set <- sapply(topo_size_all$group, function(x) strsplit(x, "_", fixed = TRUE)[[1]][2])

topo_size_all$fossils <- sapply(topo_size_all$group, function(x) strsplit(x, "_", fixed = TRUE)[[1]][3])

topo_size_all$models <- sapply(topo_size_all$group, function(x) strsplit(x, "_", fixed = TRUE)[[1]][4])

topo_size_all$lek <- substr(rownames(topo_size_all), 0 , 3)

ggplot(topo_size_all, aes(x = alignment, y = mean.c.size)) + 
  geom_boxplot(fill = viridis(10, alpha = 0.2)[4]) +
  theme(axis.text.x = element_text(size=14, angle=45, hjust = 1, vjust = 1)) + 
  labs(x = "Alignment", y = "Topologic space size") +
  facet_wrap(~ data.set + models)




# similarity
out <- lapply(pnts_lks, function(x){
  
  Y <- space_similarity(formula = chain ~ x + y, data = x, pb = FALSE, type = "proportional.overlap")
  
  Y$mean.overlap <- Y$mean.overlap -  mean(Y$mean.overlap)
  # Y$overlap <- Y$overlap / max(Y$overlap)
  
  return(Y)  
})

topo_similarity_all <- do.call(rbind, out)

topo_similarity_all2 <- topo_similarity_all[, c("group.1",  "group.2", "mean.overlap")]
names(topo_similarity_all2) <- c("group.2", "group.1", "mean.overlap")

topo_similarity_all <- rbind(topo_similarity_all[, c( "group.1", "group.2", "mean.overlap")], topo_similarity_all2[, c( "group.1", "group.2", "mean.overlap")])

topo_similarity_all$alignment <- sapply(topo_similarity_all$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][1])

topo_similarity_all$data.set <- sapply(topo_similarity_all$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][2])

topo_similarity_all$fossils <- sapply(topo_similarity_all$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][3])

topo_similarity_all$models <- sapply(topo_similarity_all$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][4])


topo_similarity_all$lek <- substr(rownames(topo_similarity_all), 0 , 3)


ggplot(topo_similarity_all, aes(x = alignment, y = mean.overlap)) + 
  geom_boxplot(fill = viridis(10, alpha = 0.2)[4]) +
  theme(axis.text.x = element_text(size=14, angle=45, hjust = 1, vjust = 1)) + 
  labs(x = "Alignment", y = "Topologic space similarity") +
  facet_wrap(~ data.set + models)

```


### Clumping molecular clocks

#### Old data set
```{r plot tree distances ggplots old all fossils clumped clocks, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("old", X$chain), ]   
  X$clock <- ifelse(grepl("global", X$chain), "Global", "Uexp")
  X$chain <- gsub("_global|_Uexp", "", X$chain)
  
    if (nrow(X) > 0)
    ggplot(data = X, aes(x = x, y = y)) +
    geom_path(alpha = 0.1, size = 0.75, show.legend = FALSE) +
    geom_point(aes(shape = clock,  col = clock), size = 4) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_color_viridis_d(begin = 0.2, end = 0.8, alpha = 0.75) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %d trees", nrow(X) / length(unique(X$chain))))

    })

ggs
```

#### New data set
```{r plot tree distances ggplots new all fossils clumped clocks, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("new", X$chain), ]   
  X$clock <- ifelse(grepl("global", X$chain), "Global", "Uexp")
  X$chain <- gsub("_global|_Uexp", "", X$chain)
  
  ggplot(data = X, aes(x = x, y = y)) +
    geom_path(alpha = 0.1, size = 0.75, show.legend = FALSE) +
    geom_point(aes(shape = clock,  col = clock), size = 4) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_color_viridis_d(begin = 0.2, end = 0.8, alpha = 0.75) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %d trees", nrow(X) / length(unique(X$chain))))

    })

ggs
```


## Early fossils using only pairwise shared tips
```{r make tree distances early fossils, eval = FALSE}

#selected leks
sel_leks <- c("SUR", "CCE", "HC1", "BR1", "TR1")
# sel_leks <- c("SUR", "TR1")

tree_names <- grep("early", names(rb.trees), value = TRUE)

pnts_lks <- lapply(sel_leks, function(i) {
  
  print(i)
  lek_trees <- grep(i, tree_names, value = TRUE)
  
  multi_list <- lapply(lek_trees, function(x) read.tree(file.path("./output/most_recent_revbayes_models/", x), keep.multi = TRUE))
  
  names(multi_list) <- sapply(basename(lek_trees), function(x) paste(strsplit(x, "_", fixed = TRUE)[[1]][1:5], collapse = "_"))
  
  pnts <- try(custom_treespace(multi_list, n.points = 100, method = "RF", cl = 10), silent = TRUE)

  if (!is(pnts, "try-error")) 
    return(pnts) else
      return(NULL)
})

names(pnts_lks) <- sel_leks

sapply(pnts_lks, is.null)

saveRDS(pnts_lks, file = "./output/topology/tree_distance_early_fossils_pairwise_shared.RDS")

```

#### New and old data sets
```{r plot tree distances ggplots early fossils, eval = TRUE, warning=FALSE}

pnts_lks <- readRDS("./output/topology/tree_distance_early_fossils_pairwise_shared.RDS")

ggs <- lapply(pnts_lks, function(X) {

    ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = viridis(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") + 
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain)))) 

    })

ggs

```

#### New data set
```{r plot tree distances ggplots new early fossils, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("new", X$chain), ]   
  
  ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = viridis(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain))))
  })

ggs


```

#### Old data set
```{r plot tree distances ggplots old early fossils, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("old", X$chain), ]   
  
  if (nrow(X) > 0)
  ggplot(data = X, aes(x = x, y = y, fill = generation)) +
    geom_path(alpha = 0.25, aes(colour = generation), size = 0.75, show.legend = FALSE) +
    scale_colour_gradient(low = "red", high = "yellow") +
    geom_point(shape = 21, size = 4, colour = "white", alpha = 0.6) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_fill_gradientn(colours = viridis(256)) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain))))
  })

ggs


```

#### Summary boxplots 
Values were mean-centered within lek

```{r, eval = TRUE, message=FALSE, warning=FALSE}

out <- lapply(pnts_lks, function(x){
  
    Y <- space_size(formula = chain ~ x + y, data = x, pb = FALSE)
  Y$mean.c.size <- Y$size - mean(Y$size)
  # Y$size <- Y$size / max(Y$size)
  
  return(Y)  
})

topo_size_early <- do.call(rbind, out)

topo_size_early$alignment <- sapply(topo_size_early$group, function(x) strsplit(x, "_", fixed = TRUE)[[1]][1])

topo_size_early$data.set <- sapply(topo_size_early$group, function(x) strsplit(x, "_", fixed = TRUE)[[1]][2])

topo_size_early$fossils <- sapply(topo_size_early$group, function(x) strsplit(x, "_", fixed = TRUE)[[1]][3])

topo_size_early$models <- sapply(topo_size_early$group, function(x) strsplit(x, "_", fixed = TRUE)[[1]][4])

topo_size_early$lek <- substr(rownames(topo_size_early), 0 , 3)

ggplot(topo_size_early, aes(x = alignment, y = mean.c.size)) + 
  geom_boxplot(fill = viridis(10, alpha = 0.2)[4]) +
  theme(axis.text.x = element_text(size=14, angle=45, hjust = 1, vjust = 1)) + 
  labs(x = "Alignment", y = "Topologic space size") +
  facet_wrap(~ data.set + models)

# similarity

out <- lapply(pnts_lks, function(x){
  
  Y <- space_similarity(formula = chain ~ x + y, data = x, pb = FALSE, type = "proportional.overlap")
  
  Y$mean.overlap <- Y$mean.overlap -  mean(Y$mean.overlap)
  # Y$overlap <- Y$overlap / max(Y$overlap)
  
  return(Y)  
})

topo_similarity_early <- do.call(rbind, out)

topo_similarity_all2 <- topo_similarity_all[, c("group.1",  "group.2", "mean.overlap")]
names(topo_similarity_all2) <- c("group.2", "group.1", "mean.overlap")

topo_similarity_all <- rbind(topo_similarity_all[, c( "group.1", "group.2", "mean.overlap")], topo_similarity_all2[, c( "group.1", "group.2", "mean.overlap")])

topo_similarity_early$alignment <- sapply(topo_similarity_early$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][1])

topo_similarity_early$data.set <- sapply(topo_similarity_early$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][2])

topo_similarity_early$fossils <- sapply(topo_similarity_early$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][3])

topo_similarity_early$models <- sapply(topo_similarity_early$group.1, function(x) strsplit(x, "_", fixed = TRUE)[[1]][4])

topo_similarity_early$lek <- substr(rownames(topo_similarity_early), 0 , 3)

ggplot(topo_similarity_early, aes(x = alignment, y = mean.overlap)) + 
  geom_boxplot(fill = viridis(10, alpha = 0.2)[4]) +
  theme(axis.text.x = element_text(size=14, angle=45, hjust = 1, vjust = 1)) + 
  labs(x = "Alignment", y = "Topologic space similarity") +
  facet_wrap(~ data.set + models)

```


### Clumping molecular clocks

#### Old data set
```{r plot tree distances ggplots old early fossils clumped clocks, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("old", X$chain), ]   
  X$clock <- ifelse(grepl("global", X$chain), "Global", "Uexp")
  X$chain <- gsub("_global|_Uexp", "", X$chain)
  
    if (nrow(X) > 0)
    ggplot(data = X, aes(x = x, y = y)) +
    geom_path(alpha = 0.1, size = 0.75, show.legend = FALSE) +
    geom_point(aes(shape = clock,  col = clock), size = 4) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_color_viridis_d(begin = 0.2, end = 0.8, alpha = 0.75) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain))))

    })

ggs
```

#### New data set
```{r plot tree distances ggplots new early fossils clumped clocks, eval = TRUE, warning=FALSE}

ggs <- lapply(pnts_lks, function(X) {
     
  X <- X[grep("new", X$chain), ]   
  X$clock <- ifelse(grepl("global", X$chain), "Global", "Uexp")
  X$chain <- gsub("_global|_Uexp", "", X$chain)
  
  ggplot(data = X, aes(x = x, y = y)) +
    geom_path(alpha = 0.1, size = 0.75, show.legend = FALSE) +
    geom_point(aes(shape = clock,  col = clock), size = 4) +
      theme(panel.background = element_blank(), axis.line = element_line(color = "grey"),
      panel.spacing = unit(0.1, "lines"), axis.title.x = element_text(vjust = -0.5), axis.title.y = element_text(vjust = 1.5)) +
    facet_wrap(~chain, nrow = 4) +
      scale_color_viridis_d(begin = 0.2, end = 0.8, alpha = 0.75) +
      labs(x = "MDS_v1", y = "MDS_v2") +
    ggtitle(label = sprintf("Tree space for %1.0f trees", nrow(X) / length(unique(X$chain))))

    })

ggs
```

# Stats

## Size of the topological space
```{r, eval = FALSE}
# combine data sets
topo_size_all$fossils <- "all"
topo_size_early$fossils <- "early"
topo_size <- rbind(topo_size_all, topo_size_early)

iterations <- 5000

priors <- c(set_prior("normal(0, 1.5)", class = "Intercept"), set_prior("normal(0, 1.5)", class = "b"))

size_mod <- brm(size ~ alignment + data.set + models + fossils, data = topo_size, prior = priors, iter = iterations, chains = 3, silent = 2, seed = 5)

```

Intercept-included model:
```{r, eval = TRUE}

size_mods <- readRDS("./data/processed/regression_results_topological_size.rds")

extended_summary(
    size_mods,
    n.posterior = 1000,
    fill =  viridis(10)[7],
    trace.palette = viridis,
    remove.intercepts = TRUE,
    highlight = TRUE,
    trace = TRUE,
    return = FALSE
)

```

Contrasts for alignment strategies:
```{r, eval = TRUE}

# contrasts
contrasts(
    fit = size_mods,
    predictor = "alignment",
    n.posterior = 2000,
    level.sep = " VS ",
    html.table = TRUE,
    plot = TRUE,
    highlight = TRUE,
    fill = fill_color
)

```


## Similarity of the topological space
```{r, eval = FALSE}
topo_similarity_all$fossils <- "all"
topo_similarity_early$fossils <- "early"

shared_columns <- intersect(names(topo_similarity_all), names(topo_similarity_early))

topo_similarity <- rbind(topo_similarity_all[, shared_columns], topo_similarity_early[, shared_columns])

similarity_mod <- brm(mean.overlap ~ alignment + data.set + models + fossils, data = topo_similarity, prior = priors, iter = iterations, chains = 3, silent = 2, seed = 5, file = "./data/processed/regression_results_topological_similiarity")

```

Intercept-included model:
```{r, eval = TRUE}

similarity_mod <- readRDS("./data/processed/regression_results_topological_similiarity.rds")


extended_summary(
    similarity_mod,
    n.posterior = 1000,
    fill =  viridis(10)[7],
    trace.palette = viridis,
    remove.intercepts = TRUE,
    highlight = TRUE,
    trace = TRUE,
    return = FALSE
)


```

Contrasts for alignment strategies:
```{r, eval = TRUE}

# contrasts
contrasts(
    fit = similarity_mod,
    predictor = "alignment",
    n.posterior = 2000,
    level.sep = " VS ",
    html.table = TRUE,
    plot = TRUE,
    highlight = TRUE,
    fill = fill_color
)



```


<font size="4">R session information</font>

```{r session info, echo=F}

sessionInfo()

```
